\emph{dfoliatR}: An R package for detection and analysis of insect defoliation signals in tree rings

### 2020-05-05 paper_elsevier/paper_elsevier.tex
Ann edits and references

  \journal{Tree-Ring Research} % Sets Journal name

@@

  \title{\emph{dfoliatR}: An R package for detection and analysis of insect
defoliation signals in tree rings}
    \author[a,b]{Christopher H. Guiterman\corref{1}}
   \ead{chguiterman@email.arizona.edu} 
    \author[a,c]{Ann M. Lynch}
   \ead{lyncha@email.arizona.edu} 
    \author[c]{Jodi N. Axelson}
   \ead{jodi.axelson@berkeley.edu} 
      \address[a]{Laboratory of Tree-Ring Research, University of Arizona, 1215 E Lowell
St.~Box 210045, Tucson, AZ, 85721}
    \address[b]{Three Pines Forest Research, LLC, PO Box 225, Etna, NH, 03750}
    \address[c]{U.S. Forest Service, Rocky Mountain Research Station, 1215 E Lowell
St.~Box 210045, Tucson, AZ, 85721}
    \address[d]{Dept of Environmental Science, Policy \& Management, University of
California, Berkeley, Berkeley, CA 94720}
      \cortext[1]{Corresponding Author}

  \begin{abstract}
  We present a new R package to provide dendroecologists with tools to
  identify, quantify, analyze, and visualize growth suppression events in
  tree rings produced by insect defoliation. The `dfoliatR' library is
  based on the Fortran V program OUTBREAK, and builds on existing
  resources in the R computing environment. `dfoliatR' expands on OUTBREAK
  to provide greater control of supression thresholds, additional output
  tables, and high-quality graphics. To use `dfoliatR' requires
  standardized ring-width measurements from insect host trees and an
  indexed tree-ring chronology from local non-host trees. It performs an
  indexing procedure to remove the climatic signal represented in the
  non-host chronology from the host-tree series. It then infers
  defoliation events in individual trees. Site-level analyses identify
  outbreak events that synchronously affect a user-defined number or
  proportion of the host trees. Functions are available for summary
  statistics and graphics of tree- and site-level series.

@@

The \texttt{dfoliatR} library adds to a growing suite \texttt{R}
packages designed for dendrochronology. Stemming from the \texttt{dplR}
library (Bunn 2008) that enables \texttt{R} users to read and write an
array of tree-ring data formats, standaridize ring widths, build and
evaluate chronologies, perform qaulity control (to name a few), one can
now also measure ring widths (Lara et al. 2015, Shi and Xiang 2019),
perform and check crossdating (Bunn 2010), and perform many analytical
tests (Zang and Biondi 2015, Jevšenak and Levanič 2018). Tools for
assessing stand dynamics and disturbance analyses are under rapid
development, with new packages for assessing growth and release events
(\texttt{TRADER}: Altman et al. 2014), metrics of growth resilience
(\texttt{pointRes}: Maaten-Theunissen et al. 2015), and fire history
(\texttt{burnr}: Malevich et al. 2018). The key objective of
\texttt{dfoliatR} is to provide tools to identify and analyse insect
defoliation and outbreak events by building on the methods employed the
FORTRAN program OUTBREAK (Swetnam and Lynch 1989). What sets
\texttt{dfoliatR} apart from packages such as \texttt{TRADER} is that it
explicitly performs an indexing procedue on host-tree series to remove
climatic and other non-defoliation related signals represented by
non-host tree species. Insect defoliation signals are identified in the
disturbance index by the duration and magnitude of negative departures.

\texttt{dfoliatR} draws upon data formats in \texttt{dplR} that are
commonly employed by other tree-ring libraries. It uses and outputs data
formats that faciliate the use packages embodied by \texttt{tidyverse}
(Wickham et al. 2019) that include efficient data manipulation
(\texttt{dplyr}: Wickham et al. 2020) and graphics (\texttt{ggplot2}:
Wickham 2016).

In this paper, we describe the statistical methods employed by
\texttt{dfoliatR}, its availability, and run through analyses for a
signle site in New Mexico. Users need not have much experience in
\texttt{R} to replicate the analyses and graphics below. All \texttt{R}
code presented below is executatable in an \texttt{R} session once the
required libraries are installed and loaded. Support documentation in
addition to this paper is provided within the package via standard help
menus (accessed by typing \texttt{?} before a function name) and on the
package website (\url{https://chguiterman.github.io/dfoliatR/}), which
includes up-to-date vignettes that describe in detail the functionality
of the software. Code to create a preprint of this manuscript including
the \texttt{R} scripts is available from
\url{https://github.com/chguiterman/dfoliatR_paper}.

@@

The \texttt{dfoliatR} library requires two sets of tree-ring data to
identify defoliation and outbreak events:

\begin{itemize}
\tightlist
\item
  Standardized ring-width series for individual trees of the host
  species
\item
  Standardized tree-ring chronology from a local non-host species
\end{itemize}

Users can develop these data sets in software of their choosing, such as
\texttt{dplR} or ARSTAN. It is important that the host-tree data include
only one tree-ring series per tree. Both \texttt{dplR} and ARSTAN have
options for averaging mutliple sample series into a tree-level series.
The tree-ring series and chronology can be read into \texttt{R} via
several available \texttt{dplR} functions.

At the hear of \texttt{dfoliatR} lies two functions:
\texttt{defoliate\_trees()} and \texttt{outbreak()}. These identify
defoliation event son individual trees and thenc omposite across
mutliple trees for a given site to identify outbreak events. It is up to
the researcher to distinguish these separate spatial scales of analysis,
\texttt{dfoliatR} then provides a set of functions to visualize and
statistically summarize tree-level defoliation and site-level outbreak
events, which will be discussed further.

@@

The \texttt{defoliate\_trees()} function will usually be used first. It
performs two processes: removing the non-defoliatation growth signal
from the host-tree series and identifying defoliation events for each
host tree. The climatic or non-defoliation signals in each host-tree
series is represented by the non-host chronology or a climate
reconstruction. This indexing procedure generates the ``growth
suppression index'' (GSI) in which disturbance is the predominant
signal. The GSI is calculated as in Swetnam et al. (1985) and Swetnam
and Lynch (1989), \begin{align}
\textrm{GSI}_{i} = \textrm{H}_{i} - \left( \textrm{NH}_{i} - \overline{\textrm{NH}} \right) \frac{\sigma_{\textrm{H}}}{\sigma_{\textrm{NH}}} \
\end{align} where H and NH are the host tree series and the non-host
chronology, in year i, respectively. Only the common period between each
host-tree series and the non-host chronology are used in Equation 1. The
non-host chronology is scaled by its mean
(\(\overline{\textrm{NH}} \approx 1.0\)) and multiplied by the ratio of
host and non-host standard deviations
(\(\frac{\sigma_{\textrm{H}}}{\sigma_{\textrm{NH}}}\)), which
approximates the variance of the host tree series. This ``corrected''
non-host chronology is subtracted from the host-tree series.

Negative departures in the normalized GSI that surpass user-defined
thresholds in duration and magnitude are defined as \emph{defoliation
events}. As in OUTBREAK, defoliation events have a single year of
maximum departure (the variable `max\_reduction' in
\texttt{defoliate\_trees()}) that is used to assess the magnitude of
growth suppression. The default setting is -1.28 (in units of standard
deviation), which was previously determined as representative of WSBW
effects (\textbf{cite}). Duration is assessed by examining sequences of
negative GSI before and after the year of maximum departure. Each
defoliation event is allowed one positive excersion on each side of the
maximum departure year. Duration is assessed across the entire sequence
that includes up to two positive excursions. The default duration is
eight years, following OUTBREAK, as is commonly used in WSBW studies
(\textbf{cite}). Defoliator insects vary in their durations of
defoliations and outbreaks, so researchers should exmaine their data and
perform sensitivity tests to to find a representative event duration.

Researchers are given the options to extend defoliation events by
bridging between successive events and allowing potentially
short-duration events that occur at the end of the series. Bridging
allows multiple events that are separated by a single year to be
combined and considered as one defolaition event. Each may have up to
five years with positive growth (\textbf{CHECK}). Series end events is
an option that can be used when the trees were sampled during a known
defoliation event, which in many cases is what led the to the site being
sampled for the reconstruction. This option eliminates the duration
parameter, but all other thresholds, and bridging, still apply. These
options are not available in OUTBREAK.

\hypertarget{identifying-outbreak-events}{%
\subsection{Identifying Outbreak
Events}\label{identifying-outbreak-events}}

The individual tree defoliation series are composited in an additional
step to identify \emph{outbreak events} that synchronously affect
mutliple trees. Users have options to define the number and/or the
proportion of trees required for a defoliation event to be considered an
outbreak.

@@

The \texttt{dfoliatR} is provided free and open source (Guiterman 2020).
It is provided to \texttt{R} users via the Comprehensive R Archive
Network (CRAN; \url{https://cran.r-project.org/}). To install
\texttt{dfoliatR} from CRAN use

@@

Issues, bug reports, and ideas for improving \texttt{dfoliatR} can be
posted to \url{https://github.com/chguiterman/dfoliatR/issues}. As an
Open Source library, we welcome and encourage community involvement in
future development. The best ways to contribute to \texttt{dfoliatR} are
through standard Github procedures or by contacting the first author.

@@

\hypertarget{tree-level-defoliation-events}{%
\subsection{Tree-level defoliation
events}\label{tree-level-defoliation-events}}

\hypertarget{site-level-events}{%
\subsection{Site-level events}\label{site-level-events}}

\hypertarget{extensions}{%
\section{Extensions}\label{extensions}}

>>>>

  \journal{Dendrochronologia} % Sets Journal name

@@

  \title{\emph{dfoliatR}: An R package for detection and analysis of insect defoliation signals in tree rings}
    \author[a,b]{Christopher H. Guiterman\corref{1}}
   \ead{chguiterman@arizona.edu} 
    \author[a,c]{Ann M. Lynch}
   \ead{ann.m.lynch@usda.gov} 
    \author[c]{Jodi N. Axelson}
   \ead{jodi.axelson@berkeley.edu} 
      \address[a]{Laboratory of Tree-Ring Research, University of Arizona, 1215 E Lowell St.~Box 210045, Tucson, AZ, 85721}
    \address[b]{Three Pines Forest Research, LLC, PO Box 225, Etna, NH, 03750}
    \address[c]{U.S. Forest Service, Rocky Mountain Research Station, 1215 E Lowell St.~Box 210045, Tucson, AZ, 85721}
    \address[d]{Dept of Environmental Science, Policy \& Management, University of California, Berkeley, Berkeley, CA 94720}
      \cortext[1]{Corresponding Author}

  \begin{abstract}
  We present a new R package to provide dendroecologists with tools to infer, quantify, analyze, and visualize growth suppression events in tree rings caused by insect defoliation, and to reconstruct defoliator outbreak chronologies. `dfoliatR' is based on the OUTBREAK program and builds on existing resources in the R and dplR computing environments. It performs an indexing procedure to remove climatic signals, represented in non-host chronologies or other reconstructed climate series, from the host-tree series. It then infers defoliation events in individual trees based on user-specified thresholds. Site-level analyses identify outbreak events that synchronously affect user-defined numbers or proportions of involved host trees. Functions are provided for summary statistics and graphics of tree- and site-level series. `dfoliatR' expands on OUTBREAK to provide greater control of suppression thresholds, additional output tables, and high-quality graphics. We provide two example data sets and script to enable users to gain familiarity with the package and its capabilities. The source code is available in the Comprehensive R Archive Network (CRAN) and on GitHub.

@@

Variation in the width and morphology of annual radial growth rings in trees permits dating and quantification of past forest insect defoliator outbreaks. Defoliation can be distinguished from climate- and other disturbance-related influences by comparing ring-width or other annually-resolved features in the wood of host species to that of non-host species or annually-resolved climate records. The effect of defoliation on radial growth of trees has been recognized since the 1860s, and used to reconstruct outbreak regimes since the 1950s (Blais 1954; Lynch 2012). It was not until the 1980s, however, that precise dendrochronological techniques were applied for inferring defoliation events and reconstructing defoliator outbreak regimes (Swetnam, Thompson, and Sutherland 1985; Speer 2010; Lynch 2012). The first studies (Swetnam, Thompson, and Sutherland 1985; Swetnam and Lynch 1989, 1993) focused on developing historical reconstructions of western spruce budworm (WSBW; \emph{Choristoneura freemani} (Lepidoptera: Tortricidae); previously known as \emph{C. occidentalis}) outbreak regimes. The methodology has since been successfully applied to a wide range of defoliator species, most of which are conifer herbivores, and has evolved in sophistication for a wide range of ecosystem situations (Lynch 2012).

The main dendrochronological tool for inferring, dating, and characterizing defoliator outbreaks from tree-ring records has been the software routine OUTBREAK (Swetnam, Thompson, and Sutherland 1985; Holmes and Swetnam 1986; Swetnam and Lynch 1989). OUTBREAK computes indices (described later in detail) of suppressed growth by subtracting a detrended and standardized climate series (a ``control'' chronology) from individual host-tree detrended and standardized radial growth series after the host and non-host series have been brought to a common variance. The non-host chronology usually consists of a site chronology developed from non-host tree species growing on a climate-sensitive site, but a gridded climate data point series, like the North American Drought Atlas (Cook and Krusic 2004) also suffices. If the host and non-host species respond similarly to climate (which can and should be tested), the derived series retains variability that the host and non-host series do not have in common, generally the insect signal and some unexplained variability (noise). The user defines a rule base specifying the magnitude and duration that a period of indexed growth suppression must meet or surpass for a period of suppressed growth to be inferred as a defoliation event at the tree level.

Though powerful, OUTBREAK is outdated and increasingly difficult to use in modern computing environments. It was written in FORTRAN V with inherently severe restrictions, as RAM and disk space were limited at that time (256 kb and 10 MB, respectively) and FORTRAN conventions imposed very strict formatting, file naming, and output conventions. The program lacks a graphical interface or capabilities, forcing users to import generated text files into spreadsheets or other software to assess results and performing analyses. Furthermore, OUTBREAK can only handle one test at a time, creating barriers to batch operation and a large burden for researchers with a datasets including multiple sites. We developed \texttt{dfoliatR} as an \texttt{R}- and \texttt{dplR}-based library to overcome these issues.

\texttt{dfoliatR} adds to a growing suite of dendrochronology packages in the \texttt{R} computing environment (R Core Team 2019). Stemming from the \texttt{dplR} library (Bunn 2008) that enables \texttt{R} users to read and write an array of tree-ring data formats, standardize ring width series, build and evaluate chronologies, and perform quality control (to name a few), one can now also measure ring widths from scanned images of prepared samples (Shi et al. 2019; Lara, Bravo, and Sierra 2015), conduct and check crossdating (Bunn 2010), analyze sub-annual anotomical features (Campelo et al. 2016; Rathgeber et al. 2011), and perform many analytical tests (Jevšenak and Levanič 2018; Zang and Biondi 2015). Tools for assessing stand dynamics and disturbance analyses are under rapid development, with new packages for assessing growth and release events (\texttt{TRADER}: Altman et al. 2014), metrics of growth resilience (\texttt{pointRes}: van der Maaten-Theunissen, van der Maaten, and Bouriaud 2015), and fire history (\texttt{burnr}: Malevich, Guiterman, and Margolis 2018). The key objective of \texttt{dfoliatR} is to provide tools to identify and analyze insect defoliation and outbreak events by building on the methods employed by OUTBREAK. It capitalizes on the robust software already available in \texttt{R} by using \texttt{dplR} data formats for incoming tree-ring series and providing output data formats embodied by the \texttt{tidyverse} (Wickham et al. 2019) that include efficient data manipulation (Wickham et al. 2020) and graphics (Wickham 2016).

In this paper, we describe the statistical methods employed by \texttt{dfoliatR}, compare results to those produced by OUTBREAK, and present an example analysis inlcuding test data sets and script. Users need not have much experience in \texttt{R} to replicate the analyses and graphics as presented. The \texttt{R} code below is executable in an \texttt{R} session once the required libraries are installed and loaded. Support documentation in addition to this paper is provided within the package via standard help menus and on the package website (\url{https://chguiterman.github.io/dfoliatR/}), which includes up-to-date vignettes that describe various routines. Code to generate a preprint of this manuscript, including the \texttt{R} scripts and tubular and graphical output is available from \url{https://github.com/chguiterman/dfoliatR_paper}.

@@

The \texttt{dfoliatR} library requires two sets of tree-ring data to infer defoliation and outbreak events:

\begin{itemize}
\tightlist
\item
  Standardized ring-width series for individual trees of the host species
\item
  A standardized tree-ring chronology from a local non-host species, or a climate reconstruction
\end{itemize}

Users can develop these data sets in the software of their choosing, such as \texttt{dplR} or ARSTAN (Cook and Holmes 1996). It is important that the host-tree data include only one tree-ring series per tree. Both \texttt{dplR} and ARSTAN have options for averaging multiple sample series into a tree-level series.

At the heart of \texttt{dfoliatR} lies two functions: \texttt{defoliate\_trees()} and \texttt{outbreak()}. These identify defoliation events on individual trees (Figure \ref{fig:fig-defol}) and then composite across multiple trees to identify outbreak events (Figure \ref{fig:fig-obr}).

@@

The \texttt{defoliate\_trees()} function is the point of entry to the \texttt{dfoliatR} library. It performs two processes, removing climate-related growth signals from the host-tree series and identifying tree-level defoliation events. The climatic or non-defoliation signals in each host-tree series are characterized by a non-host chronology or climate reconstruction. \texttt{dfoliatR} removes the non-defoliation signal by subtracting the non-host series from each host-tree series, which generates a residual index. In OUTBREAK, this residual index was termed the ``corrected index.'' We call it the ``growth suppression index'' (GSI). The GSI is calculated the same as in OUTBREAK for each host tree as
\begin{align}
\textrm{GSI}_{i} = \textrm{H}_{i} - \left( \textrm{NH}_{i} - \overline{\textrm{NH}} \right) \frac{\sigma_{\textrm{H}}}{\sigma_{\textrm{NH}}} \
\end{align}
where H and NH are the host-tree series and the non-host chronology, in year i, respectively. Only the common period between the host-tree series and the non-host chronology are used in Equation 1. The host and non-host chronologies are brought to common variance by scaling the non-host chronology by its mean (\(\overline{\textrm{NH}}\)) and multiplying by the ratio of host and non-host standard deviations (\(\frac{\sigma_{\textrm{H}}}{\sigma_{\textrm{NH}}}\)), which approximates the variance of the host tree series.

Negative departures in the normalized GSI (NGSI; GSI converted to z-scores) that surpass user-specified thresholds in duration and magnitude are defined as \emph{defoliation events}. As in OUTBREAK, NGSI must reach the magnitude threshold, assessed for each year within the departure sequence. The default setting is -1.28 (NGSI is in units of standard deviation), which was previously determined to be representative of WSBW effects (Swetnam and Lynch, 1989) and is commonly used for other species. Failure to reach the specified threshold results in sequences not being inferred as defoliation events.

Event duration is assessed by examining sequences of negative NGSI (that exceed the magnitude threshold) before and after the year of maximum departure. Each defoliation event is allowed one single-year positive excursion on each side of the year of maximum departure. Duration is computed across the entire negative sequence that allows these two positive excursions. As in OUTBREAK, the user specifies a duration threshold (minimum number of years) for a departure sequence to be inferred as a defoliation event. The default threshold is eight years, as is commonly used in WSBW studies (Swetnam and Lynch, 1989). Failure to reach the specified duration threshold results in sequences not being inferred as defoliation events, regardless of NGSI values (i.e., both thresholds must be met). Researchers can, and should, adjust the duration and magnitude parameters accordingly and critically evaluate the results, as insect species vary in the length of their outbreaks and the degree to which they can suppress tree growth.

Like OUTBREAK, users are provided an option to suspend the duration threshold at the recent end of the series in cases where an outbreak event is known to be ongoing. This should be used if the user has direct knowledge of defoliation at the site during the sampling campaign. The advantage of allowing potentially short, series-end events is that it allows the current event to be included in return-interval estimates, and can aid in identifying the start-year for the current defoliation event or outbreak.

Diverging from OUTBREAK, \texttt{dfoliatR} allows users to extend defoliation events on individual trees by bridging between sequential events. In cases where two defoliation events are separated by a single year, bridging will link them into a single event. This option was added to \texttt{dfoliatR} during the testing phase of development, when we realized that OUTBREAK deliberately omits sequential, or back-to-back events, even when these events surpass thresholds in duration and magnitude. Instead, OUTBREAK will select the one sequential event with the greatest negative departure year. In every case we assessed (described below) we felt that the OUTBREAK-omitted defoliation events should have been maintained and recorded. Due in large part to reconstructions using OUTBREAK (see papers cited by Lynch 2012), we now know considerably more about forest defoliator outbreak regimes than we did in the 1980s when OUTBREAK was written. We think that two or more prolonged events separated by a single year should in some situations be considered a single event. This is particularly relevant to WSBW and spruce budworm (\emph{C. fumiferana}), for which regimes have been reconstructed for several geographic areas and for which researchers and forest health experts have gained considerable knowledge (Schmitt, Grimble, and Searcy 1984; Sanders et al. 1985; Brookes et al. 1982, and many later publications). In many cases, the greatest growth suppression often occurs late in the outbreak due to cumulative effects accrued on a tree's resources (Swetnam and Lynch 1989, 1993). We urge caution in using the bridging option, however, because it may not be appropriate for all studied insects, such as in situations where impacted stands barely recover from one outbreak before another begins, as with pine processionary caterpillars (\emph{Thaumetopoea pityocampa} (Lepidoptera: Thaumetopoeidae)) (Carus 2004, 2009).

\hypertarget{inferring-outbreak-events}{%
\subsection{Inferring Outbreak Events}\label{inferring-outbreak-events}}

Defoliation of one or a few trees does not constitute an outbreak. To determine when defoliation becomes an \emph{outbreak event}, \texttt{dfoliatR} composites the individual tree defoliation series into a site-level chronology with the \texttt{outbreak()} function. Users have options to define the number and/or the proportion of trees required for an event to be considered an outbreak. Three parameters control whether a defoliation event constitutes an outbreak: the minimum number of trees available, the minimum number of trees recording defoliation, and the percent of trees recording defoliation. The first allows the researcher to make a judgement call as to the confidence ascribed to reduced sample depth toward the ends of their chronologies, thus compensating for the ``fading record problem'' (Swetnam, Allen, and Betancourt 1999). The second two parameters adjust the scale of defoliation considered to be an outbreak. Absolute numbers of trees and percentages can be applied separately or in conjunction, following filtering conventions in tree-ring fire history studies (Malevich, Guiterman, and Margolis 2018). We urge users to carefully consider the choice of absolute numbers in situations where the number of trees represented in the series varies with time, or the choice of percentages when sample size is small.

\hypertarget{approach}{%
\subsection{Approach}\label{approach}}

We tested \texttt{dfoliatR} against OUTBREAK by comparing NGSI to OUTBREAK's normalized corrected indices for individual trees and years, defoliation status for individual trees and years, and percentage of trees recording outbreaks at the site level, using standardized ring-width data from eight sites in British Columbia, Colorado, and New Mexico, and author-provided non-host site chronologies.

We detrended host data for both \texttt{dfoliatR} and OUTBREAK using ARSTAN (v6.1, Cook and Holmes 1996) with default double detrending (128 year wavelength and a 50\% smoothing spline). In both \texttt{dfoliatR} and OUTBREAK we used event thresholds of -1.28 normalized indices, 8 years duration, and allowed for events at the end of series in seven of eight sites. These seven sites were sampled during ongoing outbreak events (Swetnam and Lynch 1993). We found it necessary to be consistent in how we detrended and in what software (e.g., ARSTAN vs \texttt{dplR}) we employed because subtle differences in standardized ring-width indices generated between the programs transferred into differences between \texttt{dfoliatR} and OUTBREAK. In the end, we chose to only use the standardization output files from ARSTAN, which are easily read into \texttt{R} (and then \texttt{dfoliatR}) using the \texttt{dplR} package.

The \texttt{R} code to replicate our comparisons is available from \url{https://github.com/chguiterman/dfoliatR_paper}.

\hypertarget{findings}{%
\subsection{Findings}\label{findings}}

Across the 43,280 ring-width indices from 222 trees included in our evaluation, we found that \texttt{dfoliatR} and OUTBREAK compute identical growth suppression indices at 0.000 precision. We expected this outcome because both programs employ Equation 1. At the tree-level, the programs identified 11,530 years of defoliation. The programs agreed on 97.9\% of the years, leaving 927 ``difference'' years in which only one program identified defoliation on an individual tree. The differences included 102 events on 85 trees. We carefully inspected each of these events in the full context of each tree's ring-series, and categorized the differences as follows

\begin{itemize}
\item
  \emph{Series-end events} (40\% of the total) in which OUTBREAK included ``truncated outbreaks'' (for seven sites) at the end of each series. In \texttt{dfoliatR}, this option is controlled by the ``series\_end\_events'' parameter to \texttt{defoliate\_trees()}. In OUTBREAK, the option appears while changing the duration parameter (option 3). When selected, OUTBREAK will include any sequences of negative indices at the either the beginning or the end of each tree series as a defoliation event, without consideration of either duration or magnitude thresholds. In \texttt{dfoliatR}, the duration threshold is omitted and the magnitude threshold is retained in series-end-events. Each of the 13 events included in these differences did not meet the ``max\_reduction'' parameter (-1.28 NGSI) in \texttt{dfoliatR} and were excluded. In two cases, OUTBREAK included events at the beginning of the series where \texttt{dfoliatR} does not allow truncated events. In four cases, OUTBREAK omitted only the last year of the series because the index was positive. In two cases, \texttt{dfoliatR} omitted possible events because it had already included a positive NGSI excursion after the ``max\_reduction'' year, and since it will only allow one excursion on either side of the max year, the events were omitted due to short duration.
\item
  \emph{Sequential events} (36\%) in which OUTBREAK omitted events that occur one year prior or one year following an identified event. When this occurs, OUTBREAK selects the event with the greatest negative index year. On two trees, OUTBREAK omitted two of three sequential events. While inspecting these differences, we added an option to \texttt{defoliate\_trees()} that would ``bridge'' between sequential events (that surpass magnitude and duration thresholds)into single, long events. We felt that this was ecologically justified, especially for studies of WSBW, because the insect will occasionally defoliate individual trees for decades and reconstructed chronologies show outbreaks as long as \textbf{xxx (ANN CHECK)}.
\item
  \emph{Undetermined differences} (22\%) occurred in cases where OUTBREAK omitted events without clear cause that \texttt{dfoliatR} correctly identified as defoliations.
\item
  \emph{Rounding differences} (2\%) in the indices either omitted or cut short events on two trees. In both cases the indices were very close to zero, and the difference was less than the precision of the raw data measurement.
\end{itemize}

\textbf{(Bridging paragraph?)}

At the site level, OUTBREAK and \texttt{dfoliatR} produce similar time series of percent trees defoliated (Figure \ref{fig:df-obr-site}), which forms the basis for inferring outbreak occurrence, intensity, and duration. Nearly all of the differences between the programs include either more trees or a longer duration of inferred outbreak by \texttt{dfoliatR}. These arise from the inclusion of events omitted by OUTBREAK. These differences translate to greater numbers of trees in defoliation during certain events, or events recorded by single trees during periods of low sample depth.

This evaluation revealed what we believe are short-comings in how OUTBREAK identifies defoliation events on individual trees. In every one of the 102 cases we inspected, we felt that \texttt{dfoliatR} provided a more biologically and statistically appropriate assessment of defoliation. Therefore, \texttt{dfoliatR} also did a better job inferring OUTBREAK events and associated statistics at the site level.

@@

The \texttt{dfoliatR} library (Guiterman, Lynch, and Axelson 2020) is provided free and open source from the Comprehensive R Archive Network (CRAN; \url{https://cran.r-project.org/}). To install \texttt{dfoliatR} from CRAN use

@@

Issues, bug reports, and ideas for improving \texttt{dfoliatR} can be posted to \url{https://github.com/chguiterman/dfoliatR/issues}. As an Open Source library, we welcome and encourage community involvement in future development. The best ways to contribute to \texttt{dfoliatR} are through standard GitHub procedures or by contacting the first author.

@@

Once \texttt{dfoliatR} is loaded into an \texttt{R} session (via \texttt{library(dfoliatR)}) users can access two sets of tree-ring data to aid in exploring the functions, graphics, and outputs. Each data set consists of individual host-tree series and a local non-host chronology. The host-tree series were standardized using 50-year splines with a 50\% frequency response, while the non-host ring-width data were standardized using 150-year splines with a 50\% frequency response and then averaged via Tukey's biweight robust mean procedure. Host trees from Demijohn Peak (DMJ; 2902 m asl) in the San Juan Mountains of southern Colorado include Douglas-fir (\emph{Pseudotsuga menziesii}) compared against a local non-host ponderosa pine (\emph{Pinus ponderosa}) chronology (Ryerson, Swetnam, and Lynch 2003). The East Fork site (EFK; 2580 m asl) in the Jemez Mountains of northcentral New Mexico includes Douglas-fir and white fir (\emph{Abies concolor}) host trees and a ponderosa pine non-host chronology (Swetnam and Lynch 1993).

With \texttt{dfoliatR} loaded, the datasets are accessible using the \texttt{data()} function. The data object names are prefixed by their site codes. For instance, the \texttt{dmj\_*} objects come from the DMJ site and include the host-tree series (\texttt{dmj\_h}), the non-host chronology (\texttt{dmj\_nh}), the defoliation series (\texttt{dmj\_defol}), and the outbreak series (\texttt{dmj\_obr}). The same suite of data are available for EFK using the \texttt{efk\_*} prefix.

In our example scripts below, \texttt{\#\#} and \texttt{\#} denote user comments, per standard \texttt{R} coding, which are colored in brown. Text in blue denote functions; black are loaded objects, and green are quoted variables and links.

\hypertarget{tree-level-defoliation-events}{%
\subsection{Tree-Level Defoliation Events}\label{tree-level-defoliation-events}}

The function \texttt{defoliate\_trees()} performs the GSI indexing procedure on each host-tree series and then identifies defoliation events.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Load the package}
\KeywordTok{library}\NormalTok{(dfoliatR)}

\CommentTok{## Load data}
\KeywordTok{data}\NormalTok{(}\StringTok{"dmj_h"}\NormalTok{) }\CommentTok{# host tree series}
\KeywordTok{data}\NormalTok{(}\StringTok{"dmj_nh"}\NormalTok{) }\CommentTok{# non-host chronology}
\CommentTok{# View data structure for any data object via }
\CommentTok{# "View(dmj_h)" or "head(dmj_h)"}

\CommentTok{## Calculate indices, identify defoliation events}
\NormalTok{dmj_defol <-}\StringTok{ }\KeywordTok{defoliate_trees}\NormalTok{(}\DataTypeTok{host_tree =}\NormalTok{ dmj_h,}
                             \DataTypeTok{nonhost_chron =}\NormalTok{ dmj_nh,}
                             \DataTypeTok{duration_years =} \DecValTok{8}\NormalTok{,  }
                             \DataTypeTok{max_reduction =} \FloatTok{-1.28}\NormalTok{, }
                             \DataTypeTok{bridge_events =} \OtherTok{TRUE}\NormalTok{, }
                             \DataTypeTok{series_end_event =} \OtherTok{TRUE}\NormalTok{, }
                             \DataTypeTok{list_output =} \OtherTok{FALSE}\NormalTok{) }
\end{Highlighting}
\end{Shaded}

The result is long-format (stacked) data frame with five variables: ``year'', ``series'', ``gsi'', ``ngsi'', and ``defol\_status.'' The ``defol\_status'' column indicates whether that year has defoliation or not, with a set of factors that include ``nd'' for non-defoliation year, ``defol'' for a defoliation year, ``max\_defol'' for the year of maximum suppression (that acts as the basis for individual events), ``bridge\_defol'' to identify years that link subsequent events (only one is present at DMJ), and ``series\_end\_defol'' to identify defoliation at the present-end of the series.

Selecting \texttt{list\_output\ =\ TRUE} in \texttt{defoliate\_trees()} provides a list-object of data frames, each with an \texttt{rwl} object that combines the host tree and non-host series and the other columns created by \texttt{defoliate\_trees()}. This option is not used by subsequent functions in \texttt{dfoliatR}, but researchers can examine it to check the results of the GSI calculation (Equation 1), such as the non-host series after scaling to a common variance with a particular host-tree series.

The results of running \texttt{defoliate\_trees()} can be assessed through graphical and table outputs. The function \texttt{get\_defol\_events()} will provide a list of every defoliation event for every tree, with the corresponding mean ``ngsi'' value. A summary table of the results for each tree is produced by \texttt{defol\_stats()} (Table \ref{tab:tbl-defol}).

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{defol_stats}\NormalTok{(dmj_defol)}
\KeywordTok{View}\NormalTok{(dmj_defol)}
\end{Highlighting}
\end{Shaded}

The \texttt{plot\_defol()} function produces a ``ggplot'' graphics object with line segments showing the measured sequence of each series and a filled segment for each identified defoliation event (Figure \ref{fig:fig-defol}). The defoliation segments are colored by their relative severity based on their average NGSI value. The default cut-off values between ``Severe'' and ``Moderate'' is the overall mean across all events. The cut-off between ``Moderate'' and ``Minor'' is the first quartile by default. Users can re-define the breaks to suit their needs via the ``breaks'' parameter in \texttt{plot\_defol()}.

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{plot_defol}\NormalTok{(dmj_defol)}
\CommentTok{## This creates a ggplot object. Additions can }
\CommentTok{## be made to adjust plotting themes and}
\CommentTok{## aesthetics, like color.   }
\CommentTok{## For more on the features of ggplot see}
\CommentTok{## https://ggplot2.tidyverse.org/}
\CommentTok{## Adding 'scale_color_manual()' will}
\CommentTok{## change the colors that denote }
\CommentTok{## the average NGSI value for each defoliation}
\CommentTok{## event.}
\KeywordTok{library}\NormalTok{(ggplot2)}
\KeywordTok{plot_defol}\NormalTok{(dmj_defol) }\OperatorTok{+}
\StringTok{  }\KeywordTok{scale_color_manual}\NormalTok{(}\DataTypeTok{values =} \KeywordTok{c}\NormalTok{(}\StringTok{"red"}\NormalTok{, }\StringTok{"orange"}\NormalTok{, }\StringTok{"purple"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

These output functions aid in assessing the sensitivity of input parameters to \texttt{defoliate\_trees()}, including the duration and magnitude thresholds for identifying defoliation events. Using \texttt{plot\_defol()} also provides a direct assessment of the between-tree variability in defoliation.

\hypertarget{site-level-events}{%
\subsection{Site-Level Events}\label{site-level-events}}

To infer outbreak events at the site level, the function \texttt{outbreak()} composites tree-level defoliation series into a single chronology, with input parameters that control thresholds in the number and proportions of trees recording a defoliation event.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Use the defol object "dmj_defol" created above}
\NormalTok{dmj_obr <-}\StringTok{ }\KeywordTok{outbreak}\NormalTok{(dmj_defol,}
                    \DataTypeTok{filter_min_series =} \DecValTok{3}\NormalTok{,}
                    \DataTypeTok{filter_min_defol =} \DecValTok{1}\NormalTok{,}
                    \DataTypeTok{filter_perc =} \DecValTok{25}\NormalTok{)}
\end{Highlighting}
\end{Shaded}
